\plaintitle

### 2025-06-04 docs/style/article-body.tex

@@

@@

\[
Throughput_\mathrm{Mbit/s}
= \frac{DataSize_\mathrm{bits}}{ExecutionTime_\mathrm{ms}}
\times 10^{-3}
\phantom{12}(2)
\]

@@

\[
\frac{64}{24}\times10^{-3}\ \mathrm{ms}
\approx 2.667\times10^{-3}\ \mathrm{ms}
=2.667\ \mu\mathrm{s}.
\phantom{12}(3)
\]

@@

Table \ref{tab:iter0} presents the average execution time of iteration 1 on both Teensy 4.1 and Raspberry Pico Pi 2 across the input sizes. The Teensy consistently outperforms the Pico, with the gap widening as the bit size increases. These results will be used as baseline for future comparison.

@@

In addition to the results presented in Table \ref{tab:iter4}, a further tests was conducted on the Teensy using a single-loop unrolled implementation for the 64-bit size, resulting in an average execution time of \texttt{6.6626} \(\mu s\). An additional fully unrolled variant, where loops were entirely eliminated, produced a measured execution time of \texttt{0.0491} \(\mu s\). However, this measurement was later determined to be invalid due to packaging error.

@@

Table \ref{tab:iter5} presents the results of iteration 5, in which vector structure were removed in favor of fixed-size integer types \texttt{uint32\_t} and \texttt{uint64\_t}. This iteration led to an extremely efficient implementation that approaches the physical execution limits of the Teensy 4.1. Due to the removal of \texttt{std::vector}, this implementation is limited by the available integer sizes of the MCU, and could as such only be tested with at most 64 bits. The ramifications of this will be discussed in more detail in Section 7.

\begin{table}[H] \centring
\begin{tabularx}{\columnwidth}{|>{\centering\arraybackslash}X|>{\centering\arraybackslash}X|}
\hline \textbf{Data structure} & \multicolumn{1}{c|}{\textbf{Teensy ($\mu s$)}}
\\ \hline array & 0.4284 \\ unordered\_{map} & 31.5090 \\ bitset & 0.0474 \\
\hline \end{tabularx} \caption{Iteration 6 - Data type exploration}
\label{tab:iter6} \end{table}

Table \ref{tab:iter6} presents the results of iteration 6 using 64 bits input size, which was introduced to address the structural limitations encountered in iteration 5 (\emph{specifically, the fixed-width output constrains imposed by the removal of \texttt{std::vector}}). In this iteration, another set of alternative data structures were evaluated, the vector-based approach from iteration 1 serving as a baseline for comparison.

As the bitset appeared to provide gains in execution speed, we attempted to increase the input bit size to 128 bits in order to investigate whether this data structure could allow for larger outputs. However, this substantially increased the execution time to \texttt{179.8712} \(\mu s\).

@@

Observing the results, we can place them into the context of the limits imposed by the ADC, as discussed in Section 3.2. In algorithm (3) (\emph{as seen in Section 3.3}), we calculate the average execution speed required to be \(2.667 \mu s\) for the ADC that requires soldering to the microcontroller. From Tables \ref{tab:iter5} and \ref{tab:iter6} we demonstrate execution speeds well below that, with iteration 5 as well as iteration 6 -- specifically with bitsets -- being the most prominent results to discuss. Here, our implementation achieved speeds of \(0.0501 \mu s\) and \(0.0474 \mu s\) respectively which within the bounds of the ADC listed in Section 3.2. The built in ADC on Teensy has an effective sample rate of 1 MSPS (\emph{e.g.~half of MAX11102AUB}), which would require an average execution speed of \textasciitilde{}\(2.667/2 \mu s \approx 1.334 \mu s\), making our implementation suitable even for that limited conversion speed.

>>>>

\section{ABSTRACT}\label{abstract}

True random number generation is indispensable for a wide range of cryptographic and distributed computing applications, yet traditional pseudo-random number generators remain deterministic, and thus vulnerable to state‐recovery attacks. Quantum random number generation (QRNG) offers a robust alternative by harnessing inherent quantum fluctuations -- in particular, optical QRNG (OQRNG) based on shot noise provides an accessible source of entropy from which true random numbers can be generated. However, existing OQRNG solutions typically rely on expensive or bespoke hardware and defer post‐processing to host computers, limiting portability and integration. In this work, we investigate the feasibility of performing Toeplitz randomness extraction directly on resource constrained microcontrollers, thus creating a fully self‐contained QRNG system. We address two primary research areas: (RA1) developing an implementation of Toeplitz extraction that minimizes processing time on embedded hardware, and (RA2) ensuring that the extraction algorithm does not become the throughput bottleneck compared to ADC conversion or USB transfer speeds.

Our results demonstrate that, by eliminating dynamic allocations and leveraging bit‐parallel techniques, Toeplitz extraction on 64‐bit inputs can be executed in approximately \(0.05 \mu s\) on the Teensy 4.1 -- well below the \(2.667 \mu s\) threshold imposed by a 12‐bit, 2 MSPS ADC. Even accounting for variance in individual measurements, these speeds confirm that the algorithmic implementation does not constrain overall throughput. On the Pico 2, the same implementations yield sub‐microsecond performance, albeit with a wider variance. However, our most efficient implementations are limited to 64‐bit inputs -- extending to larger input bit sizes incurs significant overhead due to the absence of wider integer types in typical MCU toolchains.

In conclusion, Toeplitz extraction can be performed in real time on low‐power microcontrollers without becoming the system's performance bottleneck, validating the viability of a portable QRNG thumbstick. Future work should explore custom data types or lightweight libraries for 128- and 256-bit size extraction, as well as integration with a live OQRNG device to assess end‐to‐end randomness quality and throughput.

@@

To estimate the theoretical lower bound for execution time on the Teensy 4.1, the duration of a single CPU cycle can be calculated using the following expression:

\[
\text{Cycle time} = \frac{1}{600\, \mathrm{MHz}} \approx 1.67\, \mathrm{ns} \phantom{123}(2)
\]

Given a clock frequency of \texttt{600\ MHz}, each cycle lasts approximately 1.67 nanoseconds. Assuming a minimal implementation requires 30 instruction cycles to complete a single extraction, the theoretical execution time becomes:

\[
t_{\text{min}} = 30 \times 1.67\,\mathrm{ns} = 50.1\,\mathrm{ns} \approx 0.05\,\mu\mathrm{s} \phantom{123}(3)
\]

These theoretical limits will serve as a baseline to assess how closely our implementations approach the maximum capabilities of the Teensy 4.1 microcontroller.

@@

\[
Throughput_\mathrm{Mbit/s}
= \frac{DataSize_\mathrm{bits}}{ExecutionTime_\mathrm{ms}}
\times 10^{-3}
\phantom{12}(4)
\]

@@

\[
\frac{64}{24}\times10^{-3}\ \mathrm{ms}
\approx 2.667\times10^{-3}\ \mathrm{ms}
=2.667\ \mu\mathrm{s}.
\phantom{12}(5)
\]

@@

Table \ref{tab:iter0} presents the average execution time of the naive implementation (iteration 0) across different input sizes for both the Teensy 4.1 and the Raspberry Pi Pico 2. The results reveal that the Teensy consistently achieves lower execution times compared to the Pico, with performance differences becoming more pronounced as thje bit size increases. This behaivor aligns with the expected difference in hardware specifications, such as CPU clock frequency and memory bandwidth.

These measurements serve as the baseline for all subsequent iterations. They provide a reference point for assessing performance improvements introduced in later implementations and allow for cross-platform comparisons.

@@

In addition to the results presented in Table \ref{tab:iter4}, a further tests was conducted on the Teensy using a single-loop unrolled implementation for the 64-bit size, resulting in an average execution time of \texttt{6.6626} \(\mu s\).

An additional fully unrolled variant, where loops were entirely eliminated, produced a measured execution time of \texttt{0.0491} \(\mu s\). However, this result was later deemed invalid due to packaging error in the implementation. Specifically, the algorithm was incorrectly designed to transmit only a single output bit instead of full 32-bit expected output. This lead to an underestimation of the true execution time, as the cost of correctly formatting and sending the full result was omitted from measurement.

@@

Table \ref{tab:iter5} presents the results of iteration 5, in which vector structure were removed in favor of fixed-size integer types \texttt{uint32\_t} and \texttt{uint64\_t}. This iteration led to an extremely efficient implementation that approaches the physical execution limits of the Teensy 4.1. Due to the removal of \texttt{std::vector}, this implementation is limited by the available integer sizes of the MCU, and could as such only be tested with at most 64 bits. The ramifications of this will be discussed in more detail in Section 7. However, the results remain inconsistent---while the average is low, individual measurements vary, indicating that the implementation does not yet achieve stable execution at the architectural limit.

\vspace{1em} \begin{table}[H] \centring
\begin{tabularx}{\columnwidth}{|>{\centering\arraybackslash}X|>{\centering\arraybackslash}X|}
\hline \textbf{Data structure} & \multicolumn{1}{c|}{\textbf{Teensy ($\mu s$)}}
\\ \hline array & 0.4284 \\ unordered\_{map} & 31.5090 \\ bitset & 0.0474 \\
\hline \end{tabularx} \caption{Iteration 6 - Data type exploration}
\label{tab:iter6} \end{table}

Table \ref{tab:iter6} presents the results of iteration 6 using 64 bit size, which was introduced to address the structural limitations encountered in iteration 5. Specifically, the fixed-width output constrains imposed by the removal of vector. Several alternative data structures were evaluated, with the original vecotr-based implementation from iteration 1 serving as a baseline.

Notably, while the bitset-based implementation achieved an average execution time of 0.0474\,\(\mu s\), this value represents an average and not a consistently stable result. Individual measurements showed significant variance. This suggests that, despite nearing the theoretical limits outlined in Section 3.3, the implementation has not yet reached a consistently minimal execution time. Additional tests at 128-bit input size yielded 179.8712\,\(\mu s\), confirming that the system still operates within the physical capabilities of the Teensy 4.1.

@@

Observing the results, we can place them into the context of the limits imposed by the ADC, as discussed in Section 3.2. In equation (5) (\emph{as seen in Section 5.3}), we calculate the average execution speed required to be \(2.667 \mu s\) for the ADC that requires soldering to the microcontroller. From Tables \ref{tab:iter5} and \ref{tab:iter6} we demonstrate execution speeds well below that, with iteration 5 as well as iteration 6 -- specifically with bitsets -- being the most prominent results to discuss. Here, our implementation achieved speeds of \(0.0501 \mu s\) and \(0.0474 \mu s\) respectively which within the bounds of the ADC listed in Section 3.2. The built in ADC on Teensy has an effective sample rate of 1 MSPS (\emph{e.g.~half of MAX11102AUB}), which would require an average execution speed of \textasciitilde{}\(2.667/2 \mu s \approx 1.334 \mu s\), making our implementation suitable even for that limited conversion speed.
